---
title: "MPG comparisons between car transmission types"
author: "Kevin Tham"
date: "May 12, 2018"
output: 
  html_document:
    keep_md: true
  pdf_document: default
---

```{r setup, include=FALSE}
rm(list=ls())
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE}
if (!require("pacman"))
  install.packages("pacman", repos = "http://cran.us.r-project.org")
pacman::p_load(knitr, dplyr, ggplot2, GGally, tidyr, grid, gridExtra, car)
```

## Overview

In this report we will explore the relationship between a set of variables and miles per gallon (MPG) from a data set of a collection of cars. Specifically, we would like to answer the following two questions:

1. How different is the MPG between automatic and manual transmissions?
2. Is an automatic of manual transmission better for MPG?

Using the dataset `mtcars` we shall embark on a statistical study to address the above two questions.

## Exploratary Data Analysis

We begin the study by conducting some exploratory data analysis. First we import and examine the dataset:

```{r}
data(mtcars)
head(mtcars)
str(mtcars)
```

Some of the variables are in the wrong data type and require coercion to the correct data type: 

```{r}
mtcars$am <- factor(mtcars$am, labels = c('automatic',' manual'))
mtcars$vs <- factor(mtcars$vs, labels = c('V-shaped', 'straight'))
mtcars$cyl <- ordered(mtcars$cyl)
mtcars$gear <- ordered(mtcars$gear)
```

We can make a direct comparison between the transmission type and MPG with a boxplot:

```{r, fig.align='center', results='hold'}
ggplot(mtcars, aes(x=am,y=mpg)) +
  geom_boxplot()
```

From the boxplot we can conclude that from the dataset, cars with a manual transmission have a larger median MPG than cars with an automatic transmission. The MPG for cars with a manual transmission also appear to have a larger spread between the first and third quartiles.

In order to visualise the relationship of MPG and transmission type with the other variables we can utilise a pairplot:

```{r, fig.align='center', fig.width=11, fig.height=11, results='hold'}
ggpairs(mtcars, lower=list(combo=wrap('facethist',binwidth=0.8)))
#ggpairs(select(mtcars, mpg, cyl, disp, hp, drat, wt, qsec, am), 
#        lower=list(combo=wrap('facethist',binwidth=0.8)))
```

```{r}
fit1 <- lm(mpg ~ am, data=mtcars)
summary(fit1)

fit2 <- lm(mpg ~ ., data=mtcars[, sample(1:11)])
summary(fit2)
anova(fit2)
vif(fit2)
```